A White Noise Approach to Evolutionary Ecology

### 2020-04-02 .A White Noise Approach To Evolutionary Ecology/draft.Rmd
DaPrato&Zabczyk

---
title: "A White Noise Approach to Evolutionary Ecology"
author: "Bob Week, Steve Krone, Scott L. Nuismer, Luke J. Harmon"
date: "3/1/2020"
abstract: "We derive the dynamics of the distribution of a quantitative character and the abundance of a biological population from a stochastic partial differential equation driven by space-time white noise. In the process we develop a useful set of heuristics to operationalize the powerful, but abstract theory of white noise and measure-valued Markov processes. This approach allows us to compute the full implications of a stochastic process such as demographic stochasticity on phenotypic distributions and abundances of populations. We demonstrate the utility of our approach by deriving a model of diffuse coevolution mediated by exploitative competition for a continuum of resources. Other than trait and abundance distributions, this model predicts interaction networks parameterized by rates of interactions, competition coefficients, and selection gradients. We briefly investigate the relationship between selection gradients and competition coefficients. This illustrative investigation suggests selection gradients can be either positively or negatively correlated with competition coefficients depending on the ratio of interspecific trait variation to intraspecific trait variation. Hence, this approach can contribute to the development of a synthetic theory of evolutionary ecology by formalizing first principle derivations of dynamical equations describing populations and communities which can then be used for rigorous investigations of the relationship between feedbacks of ecological and evolutionary processes and the patterns of diversity they produce."
bibliography: ref.bib
output:
  pdf_document:
    extra_dependencies: ["float","flafter"]
    toc: false
    number_sections: true
    fig_width: 4
    fig_height: 3
    fig_caption: true
header-includes:
  - \usepackage{mathrsfs}
  - \usepackage{hyperref}
  - \usepackage[left]{lineno}
  - \usepackage{amsmath}
  - \linenumbers
editor_options: 
  chunk_output_type: console
---

@@

Reviewing these accomplishments reveals a beautiful synthesis of evolution and population ecology. However, it also reveals a gap in theoretical approaches to incorporate the intrinsically random nature of populations. Specifically, in theoretical quantitative genetics the derivation of a population's response to random genetic drift is derived in discrete time under the assumption of constant effective population size using arguments based on properties of random samples [@Lande1976]. Though this approach conveniently mimics the formalism provided by the Wright-Fisher model of population genetics, real population sizes fluctuate over time. Furthermore, since these fluctuations are themselves stochastic, it seems natural to derive expressions for the evolutionary response to demographic stochasticity and consider how the results relate to characterizations of random genetic drift. This has been done in continuous time for population genetic models without too much technical overhead, assuming a finite number of alleles. However, for populations with a continuum of types, such as a quantitative trait, this becomes a vexing mathematical challenge. Here we close this gap by combining the calculus of white noise with results on rescaled limits of branching Brownian motion processes (BBM) and stochastic partial differential equations (SPDE). Our goal has two components: 1) Establish a novel synthetic approach to theoretical evolutionary ecology that provides a formal connection between demographic stochasticity and random genetic drift in the context of quantitative traits. To show that our approach can be used to develop useful biological insights we derive a model of coevolution in an ecological network and use it to investigate the relationship between competition coefficients and selection gradients. 2) Communicate some useful properties of space-time white noise, BBM and SPDE to as wide of audience as possible. With this goal in mind we will not provide a rigorous treatment of any of these deep subjects. Instead, we introduce a set of heuristics that only require the basic concepts of Reimann integration, partial differentiation and some exposure to Brownian motion and stochastic ordinary differential equations (SDE). For a concise introduction to SDE and Brownian motion, we recommend the primer by Evans [-@9781470410544]. Rigorous treatments of SPDE and rescaled limits of BBM can be found in Walsh [-@Walsh] and Dawson [-@dawson1993measure] respectively.

To provide motivation for the stochastic equations developed later and background for our model of coevolution, we begin with \S\ref{deterministic} by briefly summarizing derivations of deterministic dynamics of biological populations. Starting with a partial differential equation (PDE), we arrive at a general set of ordinary differential equations modelling the dynamics of abundance, trait mean and trait variance. From this we observe that replacing the PDE with a SPDE provides a path to derive SDE describing the evolutionary response to demographic stochasticity. We accomplish this in \S\ref{wnc} by introducing a set of mathematical tools based on the calculus of white noise and discuss how a diffusion limit of a spatially structured branching process leads to the natural SPDE appropriate for our study. The diffusion limit in turn provides a rigorous method for constructing fitness functions used in models of evolutionary ecology. We employ these tools to derive a system of SDE generalizing our deterministic results to account for demographic stochasticity. However, although biologically insightful, these equations remain difficult to analyze and implement numerically. In \S\ref{equations} we use an assumption of normally distributed trait values to simplify these expressions into formulae that are much easier to work with. We then account for the constraint of adaptive evolution on the amount of heritable variation in a population by extending these results via a model of imperfect inheritance. The resulting equations coincide with classical results in quantitative genetics as a special case. In \S\ref{coev} we combine the derived equations of population dynamics with classical niche theory to formulate a model of coevolution across a guild of $S$ species participating in exploitative competition along a common resource continuum. In SM \S\ref{diffuse} we apply a classical theorem on rescaled limits of BBM that allow for ecological interactions to provide a rigorous derivation. To gain biological insight, in \S\ref{dynamics} we numerically integrate our model of coevolution for $S=100$ species, tracking the dynamics of traits and abundances, under scenarios of weak and strong competition. We include an account of the natural history of the simulated system and discuss the significance of demographic stochasticity for structuring ecological communities. In \S\ref{ecoevo} we provide expressions for selection gradients and competition coefficients implied by our model and use these expressions to investigate the relationship between the strengths of competition and coevolution. Finally, \S\ref{conclusion} concludes with a summary of accomplishments, a few remarks on the limits of this approach and future directions to incorporate more explicitly the genetic architecture of populations, feedbacks with ecosystem processes and the macroevolutionary history of interacting lineages.

@@

This result closely resembles Kimura's continuum-of-alleles model [@Kimura1965; @Brger1986]. The primary difference being that our model utilizes diffusion instead of convolution with an arbitrary mutation kernel. Of course, our model of mutation can be derived as an approximation to Kimura's model, which has been referred to as the Gaussian allelic approximation [in reference to the distribution of mutational effects at a given locus of a genome on the values of traits, @9780471986539], the infinitesimal model [in reference to modelling continuous traits as being encoded by an infinite number of loci each having infinitesimal effect, @Barton2017] and the Gaussian descendants approximation [in reference to offspring trait values being normally distributed around their parental values, @Turelli2017]. Alternatively, since diffusion is the continuous-time equivalent to convolution against a Gaussian kernel (SM \S\ref{diffconvequiv}), equation (\ref{p_PDE}) can also be seen as a special case of the continuum-of-alleles model.

@@

Equation (\ref{detmean}) is a continuous time analog of the well known Robertson-Price equation without transmission bias [@Robertson1966; @PRICE1970; @FRANK2012; @Queller2017; @Lion2018]. The covariance of fitness and phenotype creates change in $\bar x$ to maximize mean fitness $\bar m$. Since this change is driven by a covariance with respect to phenotypic diversity, the response in mean trait to selection is mediated by the phenotypic variance. In particular, when $\sigma^2=0$, $\bar x$ will not respond to selection. The result $\int_{-\infty}^{+\infty} x\frac{\partial^2}{\partial x^2} p(x,t)dx=0$ can be found by applying integration by parts. Following the approach taken to calculate the evolution of $\bar x$, we find the response of phenotypic variation to this model of selection and mutation is
\begin{equation}\label{vardyn_det}
\frac{d}{dt}\sigma^2(t)=\mathrm{Cov}_t\Big(m(\nu,x),(x-\bar x)^2\Big)+\mu.
\end{equation}
For the sake of space we relegate the derivation of $d\sigma^2/dt$ to SM \S\ref{var_deriv}. In the absence of mutation equation (\ref{vardyn_det}) agrees with the result derived by Lion [-@Lion2018] for discrete phenotypes. From a statistical perspective, if we think of $(x-\bar x)^2$ as a square error, then in analogy to the dynamics of the mean trait, we see that the response in $\sigma^2$ to selection can be expressed as a covariance of fitness and square error, which is defined in analogy to $\mathrm{Cov}_t(m(\nu,x),x)$. This covariance also creates change in $\sigma^2$ that maximizes mean fitness $\bar m$, which can be positive or negative depending on whether selection is stabilizing or disruptive.

@@

A notable criticism of using niche overlap to measure the intensity of competition points to cases where populations competiting on multiple niche axes exhibit overlap on at least one of the axes, but no overall niche overlap [@Holt1987]. Thus niche overlap on lower-dimensional projections of some multivariate niche space does not imply the populations compete. To illustrate with a simple example, consider two populations competing for space on the plane $\mathbb{R}^2$. If the spatial distributions of the two populations overlap, then they will overlap on both spatial axes. However, if the populations do not overlap on at least one of the spatial axes, they will have no overall spatial overlap. Furthermore, even if the species overlap on both spatial axes, they need not have any overall spatial overlap. This final result corresponds to the fact that components of niche space do not necessarily interact multiplicatively to determine the consequences for the intensity of competition. In another component of Holt's [-@Holt1987] critique, an argument is made for the potential of competition occuring without any overlap in niche space. However, this argument is based on the practical difficulty of identifying every resource axis populations are competing on and how these axes interact to determine fitness consequences. Our model avoids these caveats by assuming competition only occurs along a single dimensional resource gradient. To map the degree of niche overlap to fitness, we assume competition between individuals with niche locations $x_i$ and $x_j$ additively decreases the Malthusian fitness for the individual in species $i$ by $c_i\mathcal{O}_{ij}(x_i,x_j)$ for some $c_i\geq0$. In SM \S\ref{diffuse} we combine this niche model with equations (\ref{N}), (\ref{xbarfinal}) and (\ref{Gfinal}) to find

@@

The strong competition scenario is not quite as showy. Although the dynamics of trait means and variances tend to be far more stochastic than in the weak competition scenario, the community overall appears to quickly reach some statistical equilibrium and remain there. However, the abundances across all species in the community are very low due to strength of competition being an order of magnitude higher than in the weak competition case. Most of the species maintain abundances greater than $1000.0$, but we found one species that dropped to an abundance of about $50.0$. If we let the numerical integrator run long enough in this case, we will likely see many of the species go extinct.

@@

\begin{table}
\centering\caption{Values of model parameters used for numerical integration.}\vspace{0.2cm}
\begin{tabular}{l|l|l}
Parameter     & Description & Value                                           \\ \hline
$R$         & innate growth rate, see \S\ref{ecoevo}  & $1.0$      \\
$\theta$    & abiotic optimum & $0.0$                              \\
$a$         & strength of abiotic selection & $0.01$               \\
$c$         & strength of competition & $\{1.0\times10^{-7},5.0\times10^{-6}\}$ \\
$w$         & niche breadth & $0.1$                               \\
$U$         & total niche use & $1.0$                              \\
$\eta$      & segregation variance & $1.0$                         \\
$\mu$       & mutation rate & $1.0\times10^{-7}$                             \\
$V$ & variance of reproductive output & $5.0$
\end{tabular}\label{par_vals}
\end{table}

@@

Note that although $r_i(t)$ is referred to in the coexistence literature as the intrinsic growth rate of the population, $R_i$ is a deeper intrinsic quantity. For now we refer to $R_i$ as the _innate_ growth rate. With this connection formally established, researchers may pursue a postmodern coexistence theory that naturally includes the evolutionary dynamics of populations and the effects of demographic stochasticity.

@@

In particular, we numerically integrated system (\ref{comm_dynamics}) for $T_1=1000.0$ units of time and then continued to integrate for $T_2=1000.0$ units of time. We then calculate the covariance between the quantities $\alpha$ and $\mathfrak{C}$ across $S=100$ species for each of the last $T_2$ time steps. We assume the temporal average of these covariances across the last $T_2$ units of time approximates the expectation at equilibrium. We repeated this approach for randomly drawn $a$ and $c$ until our sample size reached $1000$. In Figure \ref{cor_coev_fig} we plot the temporally averaged values of $\mathrm{Cov}_{f_{\bar X}}(\alpha,\mathfrak{C})$ against the strength of competition $c$. Using a cubic regression, we see the correlation of coevolutionary selection gradients and competition coefficeints is negative at variance ratios below $0.5$, zero at variance ratios between $0.5$ and $1.0$, and drops below zero again at variance ratios above $1.0$.

@@

Recall that $m(\nu,x)$ is shorthand for $m(\nu(x,t),x)$. That is, $m:[0,+\infty)\times\mathbb{R}\to\mathbb{R}$. Following our assumptions of the main text, we have that $m(y,x)$ is differentiable with respect to both $x$ and $y$ and there exists $r\in\mathbb{R}$ such that $m(y,x)\leq r$ for each $y\geq0$ and $x\in\mathbb{R}$. Hence, we can apply Proposition 1.21 of Cantrell and Cosner [-@Cantrell2004] to show that $\nu(x,t)$ remains differentiable with respect to $t$ and twice differentiable with respect to $x$ for all $t\geq0$.

@@

Replacing $m$ with it's upper bound $r\in\mathbb{R}$, equation (\ref{PDE_SM}) reduces to a simple parabolic equation that can be solved using standard techniques [@stanleyfarlow1993]. In particular, when $m(\nu,x)\equiv0$ denote the solution to (\ref{PDE_SM}) by $\nu_0(x,t)$. Then, denoting
\begin{equation}
\Phi(x,t)=\frac{\exp\left(-x^2/2\mu t\right)}{\sqrt{2\pi\mu t}},
\end{equation}
we have
\begin{equation}
\nu_0(x,t)=\int_{\mathbb{R}}\Phi(x-y,t)\nu(y,0)dy.
\end{equation}
In the more general case, when $m(\nu,x)\equiv r\in\mathbb{R}$, equation (\ref{PDE_SM}) has the solution $\nu_r(x,t)=e^{rt}\nu_0(x,t)$. Hence, $\nu_r(x,t)\geq0$ for all $x\in\mathbb{R}$ and $\int_\mathbb{R}\nu_r(x,t)dx=e^{rt}N(0)<+\infty$ for all $t\geq0$. Furthermore, denoting $N_r(t)=\int_\mathbb{R}\nu_r(x,t)dx$, $p_r(x,t)=\nu_r(x,t)/N_r(t)$, $\bar x_r(t)=\int_\mathbb{R}xp_r(x,t)dx$ and $\sigma^2_r(t)=\int_\mathbb{R}(x-\bar x_r(t))^2p_r(x,t)dx$, we have
\begin{equation}
\bar x_r(t)=\int_\mathbb{R}x\int_\mathbb{R}\Phi(x-y,t)p(y,0)dydx=\int_\mathbb{R}yp(y,0)dy=\bar x(0),
\end{equation}
\begin{equation}
\sigma_r^2(t)=\int_\mathbb{R}(x-\bar x_r(t))^2\int_\mathbb{R}\Phi(x-y,t)p(y,0)dydx=\int_\mathbb{R}\Big((y-\bar x(0))^2+\mu t\Big)p(y,0)dy=\sigma^2(0)+\mu t.
\end{equation}

Hence, $|\bar x_r(t)|,\sigma_r^2(t)<+\infty$ for all $t\geq0$. For the sake of contradiction, suppose there exists $x\in\mathbb{R}$ and $t\geq0$ such that $\nu(x,t)>\nu_r(x,t)$. Then
\begin{equation}
\nu(x,t)-\nu(x,0)=\int_0^tm(\nu,x)\nu(x,s)+\frac{\mu}{2}\Delta\nu(x,s)ds>\int_0^tr\nu_r(x,s)+\frac{\mu}{2}\Delta\nu_r(x,s)ds=\nu_r(x,t)-\nu(x,0)
\end{equation}
which implies there exists $y\geq0$ and $x\in\mathbb{R}$ such that $m(y,x)>r$. But this contradicts our assumption $m(y,x)\leq r$ for all $y\geq0$ and $x\in\mathbb{R}$. So we have $\nu(x,t)\leq\nu_r(x,t)$ for each $x\in\mathbb{R}$ and $t\geq0$. This implies that $N(t)=\int_\mathbb{R}\nu(x,t)dx<+\infty$,
\begin{equation}
0\leq\int_\mathbb{R}x^2\nu(x,t)dx\leq\int_\mathbb{R}x^2\nu_r(x,t)dx<+\infty
\end{equation}
and in particular
\begin{equation}
0\leq\sigma^2(t)+\bar x^2(t)=\frac{1}{N(t)}\int_\mathbb{R}x^2\nu(x,t)dx<+\infty
\end{equation}
for each $t\geq0$.

@@

Let $g:\mathbb{R}^d\to\mathbb{R}$ be smooth. Consider the deterministic Cauchy problem

\begin{equation}\label{heateqn}
\left\{\begin{matrix}
\dot f(x,t)=&\Delta f(x,t), & (x,t)\in\mathbb{R}^d\times(0,\infty)\\
f(x,t)=&g(x), & (x,t)\in\mathbb{R}^d\times\{0\}.
\end{matrix}\right.  \tag{SM1.1}
\end{equation}

According to Evans [-@lawrenceevans2010], the fundamental solution of (\ref{heateqn}) is

\begin{equation}
\Phi(x,t)=\frac{1}{(4\pi t)^{d/2}}\exp\left(-\frac{|x|^2}{4t}\right), \ (x,t)\in(0,\infty)\times\mathbb{R}^d, \tag{SM1.2}
\end{equation}

where $|x|=\sqrt{\sum_ix_i^2}$. The solution $f(x,t)$ of PDE (\ref{heateqn}) is then given by the convolution

\begin{equation}
f(x,t)=\int_{\mathbb{R}^d}\Phi(x-y,t)g(y)dy, \ (x,t)\in(0,\infty)\times\mathbb{R}^d. \tag{SM1.3}
\end{equation}

Hence, by the fundamental theorem of calculus,

\begin{equation}
f(x,t)+\int_t^{t+1}\dot f(x,s)ds=f(x,t+1) \tag{SM1.4}
\end{equation}

>>>>

---
title: "A White Noise Approach to Evolutionary Ecology"
author: "Bob Week, Steve Krone, Scott L. Nuismer, Luke J. Harmon"
date: "3/1/2020"
abstract: "We derive the dynamics of the distribution of a quantitative character and the abundance of a biological population from a stochastic partial differential equation driven by space-time white noise. In the process we develop a useful set of heuristics to operationalize the powerful, but abstract theory of white noise and measure-valued Markov processes. This approach allows us to compute the full implications of a stochastic process such as demographic stochasticity on phenotypic distributions and abundances of populations. We demonstrate the utility of our approach by deriving a model of diffuse coevolution mediated by exploitative competition for a continuum of resources. Other than trait and abundance distributions, this model predicts interaction networks parameterized by rates of interactions, competition coefficients, and selection gradients. We briefly investigate the relationship between selection gradients and competition coefficients. This illustrative investigation suggests selection gradients can be either positively or negatively correlated with competition coefficients depending on the ratio of interspecific trait variation to intraspecific trait variation. Hence, this approach can contribute to the development of a synthetic theory of evolutionary ecology by formalizing first principle derivations of dynamical equations describing populations and communities which can then be used for rigorous investigations of the relationship between feedbacks of ecological and evolutionary processes and the patterns of diversity they produce."
bibliography: ref.bib
output:
  pdf_document:
    extra_dependencies: ["float","flafter"]
    toc: false
    number_sections: true
    fig_width: 4
    fig_height: 3
    fig_caption: true
header-includes:
  - \usepackage{mathtools}
  - \usepackage{mathrsfs}
  - \usepackage{hyperref}
  - \usepackage[left]{lineno}
  - \usepackage{amsmath}
  - \linenumbers
editor_options: 
  chunk_output_type: console
---

@@

Reviewing these accomplishments reveals a beautiful synthesis of evolution and population ecology. However, it also reveals a gap in theoretical approaches to incorporate the intrinsically random nature of populations. Specifically, in theoretical quantitative genetics the derivation of a population's response to random genetic drift is derived in discrete time under the assumption of constant effective population size using arguments based on properties of random samples [@Lande1976]. Though this approach conveniently mimics the formalism provided by the Wright-Fisher model of population genetics, real population sizes fluctuate over time. Furthermore, since these fluctuations are themselves stochastic, it seems natural to derive expressions for the evolutionary response to demographic stochasticity and consider how the results relate to characterizations of random genetic drift. This can be done in continuous time for population genetic models without too much technical overhead, assuming a finite number of alleles. However, for populations with a continuum of types, such as a quantitative trait, this becomes a vexing mathematical challenge. Here we close this gap by combining the calculus of white noise with results on rescaled limits of branching Brownian motion processes (BBM) and stochastic partial differential equations (SPDE). Our goal has two components: 1) Establish a novel synthetic approach to theoretical evolutionary ecology that provides a formal connection between demographic stochasticity and random genetic drift in the context of quantitative traits. To show that our approach can be used to develop useful biological insights we derive a model of coevolution in an ecological network and use it to investigate the relationship between competition coefficients and selection gradients. 2) Communicate some useful properties of space-time white noise, BBM and SPDE to as wide of audience as possible. With this goal in mind we will not provide a rigorous treatment of any of these deep subjects. Instead, we introduce a set of heuristics that only require the basic concepts of Reimann integration, partial differentiation and some exposure to Brownian motion and stochastic ordinary differential equations (SDE). For a concise introduction to SDE and Brownian motion, we recommend the primer by Evans [-@9781470410544]. Rigorous treatments of SPDE and rescaled limits of BBM can be found in Walsh [-@Walsh] and Dawson [-@dawson1993measure] respectively.

To provide motivation for the stochastic equations developed later and background for our model of coevolution, we begin with \S\ref{deterministic} by briefly summarizing derivations of deterministic dynamics of biological populations. Starting with a partial differential equation (PDE), we arrive at a general set of ordinary differential equations modelling the dynamics of abundance, trait mean and trait variance. From this we observe that replacing the PDE with a SPDE provides a path to derive SDE describing the evolutionary response to demographic stochasticity. We accomplish this in \S\ref{wnc} by introducing a set of mathematical tools based on the calculus of white noise and discuss how a diffusion limit of a spatially structured branching process leads to the natural SPDE appropriate for our study. The diffusion limit in turn provides a rigorous method for constructing fitness functions used in models of evolutionary ecology. We employ these tools to derive a system of SDE generalizing our deterministic results to account for demographic stochasticity. However, although biologically insightful, these equations remain difficult to analyze and implement numerically. In \S\ref{equations} we use an assumption of normally distributed trait values to simplify these expressions into formulae that are much easier to work with. We then account for the constraint of adaptive evolution on the amount of heritable variation in a population by extending these results via a model of imperfect inheritance. The resulting equations coincide with classical results in quantitative genetics as a special case. In \S\ref{coev} we combine the derived equations of population dynamics with classical niche theory to formulate a model of coevolution across a guild of $S$ species participating in exploitative competition along a common resource continuum. In SM \S\ref{diffuse} we apply a classical theorem on rescaled limits of BBM that allow for ecological interactions to provide a rigorous derivation. To gain biological insight, in \S\ref{dynamics} we numerically integrate our model of coevolution for $S=100$ species, tracking the dynamics of traits and abundances, under scenarios of weak and strong competition. We include an account of the natural history of the simulated system and discuss the significance of demographic stochasticity for structuring ecological communities. In \S\ref{ecoevo} we provide expressions for selection gradients and competition coefficients implied by our model and use these expressions to investigate the relationship between the degree of competition and coevolution. Finally, \S\ref{conclusion} concludes with a summary of accomplishments, a few remarks on the limits of this approach and future directions to incorporate more explicitly the genetic architecture of populations, feedbacks with ecosystem processes and the macroevolutionary history of interacting lineages.

@@

This result closely resembles Kimura's continuum-of-alleles model [@Kimura1965; @Brger1986]. The primary difference being that our model utilizes diffusion instead of convolution with an arbitrary mutation kernel. Of course, our model of mutation can be derived as an approximation to Kimura's model, which has been referred to as the Gaussian allelic approximation [in reference to the distribution of mutational effects at a given locus of a genome on the values of traits, @Lande1975; but see also @9780471986539], the infinitesimal model [in reference to modelling continuous traits as being encoded by an infinite number of loci each having infinitesimal effect, @Barton2017] and the Gaussian descendants approximation [in reference to offspring trait values being normally distributed around their parental values, @Turelli2017]. Alternatively, since diffusion is the continuous-time equivalent to convolution against a Gaussian kernel (SM \S\ref{diffconvequiv}), equation (\ref{p_PDE}) can also be seen as a special case of the continuum-of-alleles model.

@@

Equation (\ref{detmean}) is a continuous time analog of the well known Robertson-Price equation without transmission bias [@Robertson1966; @PRICE1970; @FRANK2012; @Queller2017; @Lion2018]. Whether or not the covariance of fitness and phenotype creates change in $\bar x$ to maximize mean fitness $\bar m$ depends on the degree to which selection is frequency dependent [@Lande1976]. Since this change is driven by a covariance with respect to phenotypic diversity, the response in mean trait to selection is mediated by the phenotypic variance. In particular, when $\sigma^2=0$, $\bar x$ will not respond to selection. The result $\int_{-\infty}^{+\infty} x\frac{\partial^2}{\partial x^2} p(x,t)dx=0$ can be found by applying integration by parts. Following the approach taken to calculate the evolution of $\bar x$, we find the response of phenotypic variation to this model of selection and mutation is
\begin{equation}\label{vardyn_det}
\frac{d}{dt}\sigma^2(t)=\mathrm{Cov}_t\Big(m(\nu,x),(x-\bar x)^2\Big)+\mu.
\end{equation}
For the sake of space we relegate the derivation of $d\sigma^2/dt$ to SM \S\ref{var_deriv}. In the absence of mutation equation (\ref{vardyn_det}) agrees with the result derived by Lion [-@Lion2018] for discrete phenotypes. From a statistical perspective, if we think of $(x-\bar x)^2$ as a square error, then in analogy to the dynamics of the mean trait, we see that the response in $\sigma^2$ to selection can be expressed as a covariance of fitness and square error, which is defined in analogy to $\mathrm{Cov}_t(m(\nu,x),x)$. Just as for the evolution of $\bar x(t)$, this covariance also creates change in $\sigma^2$ that can either increase or decrease mean fitness $\bar m$, depending on whether or not individual fitness depends on the distribution of traits across the entire population. The effect of selection on phenotypic variance can be positive or negative depending on whether selection is stabilizing or disruptive.

@@

A notable criticism of using niche overlap to measure the intensity of competition points to cases where populations competiting on multiple niche axes exhibit overlap on at least one of the axes, but no overall niche overlap [@Holt1987]. Thus niche overlap on lower-dimensional projections of some multivariate niche space does not imply the populations compete. To illustrate with a simple example, consider two populations competing for space on the plane $\mathbb{R}^2$. If the spatial distributions of the two populations overlap, then they will overlap on both spatial axes. However, if the populations do not overlap on at least one of the spatial axes, they will have no overall spatial overlap. Furthermore, even if the species overlap on both spatial axes, they need not have any overall spatial overlap. This final result corresponds to the fact that components of niche space do not necessarily interact multiplicatively to determine the consequences for the intensity of competition. In another component of Holt's [-@Holt1987] critique, an argument is made for the potential of competition occuring without any overlap in niche space. However, this argument is based on the practical difficulty of identifying every resource axis populations are competing on and how these axes interact to determine fitness consequences. Our model avoids these caveats by assuming competition only occurs along a single dimensional resource gradient. To map the degree of niche overlap to fitness, we assume competition between individuals with niche locations $x_i$ and $x_j$ additively decreases the Malthusian fitness for the individual in species $i$ by $c_i\mathcal{O}_{ij}(x_i,x_j)$ for some $c_i\geq0$. We refer to $c_i$ as the sensitivity to competition for species $i$. In SM \S\ref{diffuse} we combine this niche model with equations (\ref{N}), (\ref{xbarfinal}) and (\ref{Gfinal}) to find

@@

The strong competition scenario is not quite as showy. Although the dynamics of trait means and variances tend to be far more stochastic than in the weak competition scenario, the community overall appears to quickly reach some statistical equilibrium and remain there. However, the abundances across all species in the community are very low due to competition sensitivies being an order of magnitude higher than in the weak competition case. Most of the species maintain abundances greater than $1000.0$, but we found one species that dropped to an abundance of about $50.0$. If we let the numerical integrator run long enough in this case, we will likely see many of the species go extinct.

@@

\begin{table}
\centering\caption{Values of model parameters used for numerical integration.}\vspace{0.2cm}
\begin{tabular}{l|l|l}
Parameter     & Description & Value                                           \\ \hline
$R$         & innate growth rate, see \S\ref{ecoevo}  & $1.0$      \\
$\theta$    & abiotic optimum & $0.0$                              \\
$a$         & strength of abiotic selection & $0.01$               \\
$c$         & sensitivity to competition & $\{1.0\times10^{-7},5.0\times10^{-6}\}$ \\
$w$         & niche breadth & $0.1$                               \\
$U$         & total niche use & $1.0$                              \\
$\eta$      & segregation variance & $1.0$                         \\
$\mu$       & mutation rate & $1.0\times10^{-7}$                             \\
$V$ & variance of reproductive output & $5.0$
\end{tabular}\label{par_vals}
\end{table}

@@

Note that although $r_i(t)$ is referred to in the coexistence literature as the intrinsic growth rate of the population, $R_i$ is a deeper intrinsic quantity. For now we refer to $R_i$ as the _innate_ growth rate. Previous work has shown the importance of demographic stochasticity [@Schreiber2017] and evolutionary dynamics [@Schreiber2018] in determining ceoxistence of competing species. However, these studies consider evolution and stochasticity separately. With the connection outlined in this section formally established, researchers may pursue a postmodern coexistence theory that naturally includes both the evolutionary dynamics of populations and the effects of demographic stochasticity in a simple synthetic framework.

@@

In particular, we numerically integrated system (\ref{comm_dynamics}) for $T_1=1000.0$ units of time and then continued to integrate for $T_2=1000.0$ units of time. We then calculate the covariance between the quantities $\alpha$ and $\mathfrak{C}$ across $S=100$ species for each of the last $T_2$ time steps. We assume the temporal average of these covariances across the last $T_2$ units of time approximates the expectation at equilibrium. We repeated this approach for randomly drawn $a$ and $c$ until our sample size reached $1000$. In Figure \ref{cor_coev_fig} we plot the temporally averaged values of $\mathrm{Cov}_{f_{\bar X}}(\alpha,\mathfrak{C})$ against the sensitivity to competition $c$. Using a cubic regression, we see the correlation of coevolutionary selection gradients and competition coefficeints is negative at variance ratios below $0.5$, zero at variance ratios between $0.5$ and $1.0$, and drops below zero again at variance ratios above $1.0$.

@@

Recall that $m(\nu,x)$ is shorthand for $m(\nu(x,t),x)$. That is, $m:[0,+\infty)\times\mathbb{R}\to\mathbb{R}$. Following our assumptions of the main text, we have that $m(y,x)$ is differentiable with respect to both $x$ and $y$ and there exists $r\in\mathbb{R}$ such that $m(y,x)\leq r$ across all $y\geq0$ and $x\in\mathbb{R}$. Hence, we can apply Proposition 1.21 of Cantrell and Cosner [-@Cantrell2004] to show that $\nu(x,t)$ remains differentiable with respect to $t$ and twice differentiable with respect to $x$ for all $t\geq0$.

@@

Replacing $m$ with it's upper bound $r\in\mathbb{R}$, equation (\ref{PDE_SM}) reduces to a simple parabolic equation that can be solved using standard techniques [@stanleyfarlow1993]. In particular, when $m(\nu,x)\equiv0$ denote the solution to (\ref{PDE_SM}) by $\nu_0(x,t)$. Then, denoting
\begin{equation}
\Phi(x,t)=\frac{\exp\left(-x^2/2\mu t\right)}{\sqrt{2\pi\mu t}},
\end{equation}
we have
\begin{equation}
\nu_0(x,t)=\int_{\mathbb{R}}\Phi(x-y,t)\nu(y,0)dy.
\end{equation}
In the more general case, when $m(\nu,x)\equiv r\in\mathbb{R}$, equation (\ref{PDE_SM}) has the solution $\nu_r(x,t)=e^{rt}\nu_0(x,t)$. Hence, $\nu_r(x,t)\geq0$ for all $x\in\mathbb{R}$ and $\int_\mathbb{R}\nu_r(x,t)dx=e^{rt}N(0)<+\infty$ for all $t\geq0$. Furthermore, denoting 
\begin{equation}
\begin{matrix}
N_r(t)&=&\int_\mathbb{R}\nu_r(x,t)dx, \\
p_r(x,t)&=&\nu_r(x,t)/N_r(t), \\
\bar x_r(t)&=&\int_\mathbb{R}xp_r(x,t)dx, \\
\sigma^2_r(t)&=&\int_\mathbb{R}(x-\bar x_r(t))^2p_r(x,t)dx, 
\end{matrix}
\end{equation}
we have
\begin{equation}
\bar x_r(t)=\int_\mathbb{R}x\int_\mathbb{R}\Phi(x-y,t)p(y,0)dydx=\int_\mathbb{R}yp(y,0)dy=\bar x(0),
\end{equation}
\begin{equation}
\sigma_r^2(t)=\int_\mathbb{R}(x-\bar x_r(t))^2\int_\mathbb{R}\Phi(x-y,t)p(y,0)dydx=\int_\mathbb{R}\Big((y-\bar x(0))^2+\mu t\Big)p(y,0)dy=\sigma^2(0)+\mu t.
\end{equation}
Hence, $|\bar x_r(t)|,\sigma_r^2(t)<+\infty$ for all $t\geq0$. For the sake of contradiction, suppose there exists $x\in\mathbb{R}$ and $t\geq0$ such that $\nu(x,t)>\nu_r(x,t)$. Then
\begin{equation}
\nu(x,t)-\nu(x,0)=\int_0^tm(\nu,x)\nu(x,s)+\frac{\mu}{2}\Delta\nu(x,s)ds>\int_0^tr\nu_r(x,s)+\frac{\mu}{2}\Delta\nu_r(x,s)ds=\nu_r(x,t)-\nu(x,0)
\end{equation}
which implies there exists $y\geq0$ and $x\in\mathbb{R}$ such that $m(y,x)>r$. But this contradicts our assumption $m(y,x)\leq r$ for all $y\geq0$ and $x\in\mathbb{R}$. So we have $\nu(x,t)\leq\nu_r(x,t)$ for each $x\in\mathbb{R}$ and $t\geq0$. This implies that $N(t)=\int_\mathbb{R}\nu(x,t)dx<+\infty$,
\begin{equation}
0\leq\int_\mathbb{R}x^2\nu(x,t)dx\leq\int_\mathbb{R}x^2\nu_r(x,t)dx<+\infty
\end{equation}
and in particular
\begin{equation}
0\leq\sigma^2(t)+\bar x^2(t)=\frac{1}{N(t)}\int_\mathbb{R}x^2\nu(x,t)dx<+\infty
\end{equation}
for each $t\geq0$.

@@

For continuous $g:\mathbb{R}^d\to\mathbb{R}$, consider the deterministic Cauchy problem
\begin{equation}\label{heateqn}
\left\{\begin{matrix}
\dot f(x,t)=&\Delta f(x,t), & (x,t)\in\mathbb{R}^d\times(0,\infty)\\
f(x,t)=&g(x), & (x,t)\in\mathbb{R}^d\times\{0\}.
\end{matrix}\right.
\end{equation}
According to Evans [-@lawrenceevans2010], the fundamental solution of (\ref{heateqn}) is
\begin{equation}
\Phi(x,t)=\frac{1}{(4\pi t)^{d/2}}\exp\left(-\frac{|x|^2}{4t}\right), \ (x,t)\in(0,\infty)\times\mathbb{R}^d,
\end{equation}
where $|x|=\sqrt{\sum_ix_i^2}$. The solution $f(x,t)$ of PDE (\ref{heateqn}) is then given by the convolution
\begin{equation}
f(x,t)=\int_{\mathbb{R}^d}\Phi(x-y,t)g(y)dy, \ (x,t)\in(0,\infty)\times\mathbb{R}^d.
\end{equation}
Hence, by the fundamental theorem of calculus,
